Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-07-29 Codes/D1_Analysis.rmd
analysis updated

#--- Data used for regression analysis ---#
reg_data <-
  prepare_reg_data(
    data = final_data,
    sat_thld_m = 9, # saturated thickness has to be at least 12 meters
    ir_share_thld = 0.75, # at least 75% of total county area has to be overlapped with HPA
    balance_thld = c(-1200, 200),
    sat_cat_num = 3,
    state_ls =
      list(
        corn = c("CO", "KS", "NE", "NM", "SD", "TX", "WY"),
        soy = c("KS", "NE", "TX")
      )
  ) %>%
  data.table() %>%
  setnames("crop_type", "crop") %>%
  #--- further necessary transformation ---#
  dplyr::mutate(data = purrr::map(
    reg_data_y,
    \(reg_data_y) {
      copy(reg_data_y) %>%
        #--- bu/acre to tonne/ha ---#
        .[, yield := yield * 12.553 / 200] %>%
        .[, dry_or_not := fifelse(sat_cat == "dryland", 1, 0)] %>%
        .[, sc_dry := paste0(sc_code, "_", dry_or_not)] %>%
        .[, state_year := paste0(state_name, "_", year)] %>%
        .[, balance := -balance]
    }
  ))

@@

```{r}
set.seed(283397)

#--- global parameters ---#
sc_base <- "31011"
sc_dry_base <- "31011_0"

@@

```{r}
#++++++++++++++++++++++++++++++++++++
#+ Bootstrap
#++++++++++++++++++++++++++++++++++++
boot_results_added <-
  main_analysis %>%
  #--- bootsrap data ---#
  dplyr::mutate(boot_data = list(
    replicate(1000, boot(data), simplify = FALSE)
  )) %>%
  #--- get share data ---#
  dplyr::mutate(share_boot_data = list(
    lapply(
      1:length(boot_data),
      \(x) {
        boot_data[[x]] %>%
          .[ir == "ir" & ir_area_ratio >= 0.75, ] %>%
          .[, balance_avg := mean(balance), by = sc_code]
      }
    )
  )) %>%
  #--- yield analysis ---#
  dplyr::mutate(yield_pred_data_boot = list(
    parallel::mclapply(
      1:length(boot_data),
      \(x) {
        print(x)
        yield_analysis(boot_data[[x]], balance_seq, sat_seq_eval, sat_breaks, sat_text_data)
      },
      mc.cores = parallel::detectCores() - 2
    )
  )) %>%
  #--- share analysis ---#
  dplyr::mutate(share_pred_data_boot = list(
    parallel::mclapply(
      1:length(share_boot_data),
      \(x) {
        print(x)
        share_analysis(share_boot_data[[x]], sat_seq, sandtotal_med, silttotal_med, awc_med)
      },
      mc.cores = parallel::detectCores() - 2
    )
  )) %>%
  #--- average yield analysis ---#
  dplyr::mutate(avg_yield_pred_data_boot = list(
    parallel::mclapply(
      1:length(boot_data),
      \(x) {
        get_average_yield(yield_pred_data_boot[[x]], share_pred_data_boot[[x]])
      },
      mc.cores = parallel::detectCores() - 2
    )
  )) %>%
  dplyr::mutate(
    yield_pred_data_boot = list(rbindlist(yield_pred_data_boot)),
    share_pred_data_boot = list(rbindlist(share_pred_data_boot)),
    avg_yield_pred_data_boot = list(rbindlist(avg_yield_pred_data_boot))
  ) %>%
  dplyr::select(
    crop, data, sat_seq, sat_text_data,
    yield_pred_data, yield_pred_data_boot,
    share_pred_data, share_pred_data_boot,
    avg_yield_pred_data, avg_yield_pred_data_boot,
  )

saveRDS(boot_results_added, "Shared/Results/boot_results_added.rds")
```

```{r}
#++++++++++++++++++++++++++++++++++++
#+ Summarize
#++++++++++++++++++++++++++++++++++++
boot_results_added <- readRDS("Shared/Results/boot_results_added.rds")

>>>>

#--- Data used for regression analysis ---#
reg_data <-
  prepare_reg_data(
    data = final_data,
    sat_thld_m = 9, # saturated thickness has to be at least 12 meters
    ir_share_thld = 0.75, # at least 75% of total county area has to be overlapped with HPA
    balance_thld = c(-1200, 200),
    sat_cat_num = 3,
    state_ls =
      list(
        corn = c("CO", "KS", "NE", "NM", "SD", "TX", "WY"),
        soy = c("KS", "NE", "TX")
      )
  ) %>%
  data.table() %>%
  setnames("crop_type", "crop")

@@

```{r}

#--- global parameters ---#

@@

```{r}
set.seed(837943)

boot_results_added$yield_pred_data_boot[[1]] %>% 
  rbindlist() %>%
  .[, .(
      # y_hat_boot = median(y_hat, na.rm = TRUE),
      y_hat_se = sd(y_hat, na.rm = TRUE),
      # dif_y_hat_boot = mean(dif_y_hat, na.rm = TRUE),
      dif_y_hat_se = sd(dif_y_hat, na.rm = TRUE)
    ),
    by = .(balance, sat, sat_cat)
    ]

boot_results_added <-
  main_analysis[1, ]  %>%
  #--- bootsrap data ---#
  dplyr::mutate(boot_data = list(
    replicate(1000, boot(data), simplify = FALSE)
  )) %>%
  #--- get share data ---#
  dplyr::mutate(share_boot_data = list(
    lapply(
      1:length(boot_data),
      \(x) {
        boot_data[[x]] %>%
          .[ir == "ir" & ir_area_ratio >= 0.75, ] %>%
          .[, balance_avg := mean(balance), by = sc_code]
      }
    )
  )) %>%
  #--- yield analysis ---#
  dplyr::mutate(yield_pred_data_boot = list(
    parallel::mclapply(
      1:length(boot_data),
      \(x) {
        print(x)
        yield_analysis_manual(boot_data[[x]], balance_seq, sat_seq_eval, sat_breaks, sat_text_data)
      },
      mc.cores = parallel::detectCores() - 2
    )
  ))

```
```{r}
#++++++++++++++++++++++++++++++++++++
#+ Bootstrap
#++++++++++++++++++++++++++++++++++++
set.seed(837943)

balance_seq <- main_analysis$balance_seq[[1]]
  balance_seq <- main_analysis$balance_seq[[1]]
  sat_seq_eval <- main_analysis$sat_seq_eval[[1]]
  sat_breaks <- main_analysis$sat_breaks[[1]]
  sat_text_data <- main_analysis$sat_text_data[[1]]

temp <- lapply(1:500, get_coef) %>% rbindlist

main_analysis$data[[1]][sc_dry == "56031_1", ]

get_coef <- function(i) {

yield_data <- main_analysis$boot_data[[1]][[i]]
semi_res_corn <-
    semi_ols(
      semi_formula = semi_formula,
      c_formula = paste0("i(sat_cat, ref = 'dryland') + i(year, ref = 2009) + i(sc_dry, ref = '", sc_dry_base, "')"),
      # fe = "",
      cluster = "sc_dry",
      data = yield_data
    )

y_res_int <- semi_res_corn$fe_res

return_data[, mean(y_hat), by = sat_cat]
yield_data[, mean(yield), by = sat_cat]

return(
tidy(y_res_int) %>%
data.table() %>%
.[term %in% c("sat_cat::[9.01,23.1]", "sat_cat::(23.1,56.8]", "sat_cat::(56.8,216]"), estimate]
)

}

main_analysis <-
  main_analysis %>%
  #--- bootsrap data ---#
  dplyr::mutate(boot_data = list(
    replicate(1000, boot(data), simplify = FALSE)
  )) %>%
  #--- get share data ---#
  dplyr::mutate(share_boot_data = list(
    lapply(
      1:length(boot_data),
      \(x) {
        boot_data[[x]] %>%
          .[ir == "ir" & ir_area_ratio >= 0.75, ] %>%
          .[, balance_avg := mean(balance), by = sc_code]
      }
    )
  ))
```

```{r}
boot_data <- boot_results_added$boot_data[[1]]

balance_seq <- boot_results_added$balance_seq[[1]]
sat_seq_eval <- boot_results_added$sat_seq_eval[[1]]
sat_breaks <- boot_results_added$sat_breaks[[1]]
sat_text_data <- boot_results_added$sat_text_data[[1]]

temp <- yield_analysis(boot_data[[200]], balance_seq, sat_seq_eval, sat_breaks, sat_text_data)

ggplot(temp[sat == 50, ]) +
  geom_point(aes(y = y_hat, x = balance)) +
  ylim(0, 15)

all_results$yield_pred_data_boot[[1]][sat == 50, sd(y_hat), by = balance]

all_results$yield_pred_data_boot[[1]][sat == 10, ]

all_results$yield_pred_data_boot[[1]][sat == 10, ] %>%
ggplot(.) +
  geom_point(aes(y = y_hat, x = balance))

boot_corn <- 
  boot_results_added$yield_pred_data_boot[[1]] %>%
  rbindlist(idcol = "boot") %>%
  .[, mean(y_hat), by = .(boot, sat)]

boot_ids <- boot_corn[V1 < 11 & sat == 10, boot]

boot_results_added$boot_data[[1]][[11]] %>%
  .[, .N, by = sat_cat]

ggplot(data = boot_results_added$boot_data[[1]][[12]]) +
  geom_point(aes(y = yield, x = balance)) +
  facet_grid(sat_cat ~ .)

boot_corn[sat %in% c(10, 50), ] %>%
  dcast(boot ~ sat, value.var = "V1") %>%
  ggplot() + 
  geom_point(aes(y = `50`, x = `10`))

  ggplot(boot_corn) +
    geom_histogram(aes(x = V1)) +
    facet_grid(sat ~ .)

Check if the average yield and predicted average yield are similar.

```
```{r}
#++++++++++++++++++++++++++++++++++++
#+ Summarize
#++++++++++++++++++++++++++++++++++++
boot_results_added <- readRDS("Shared/Results/boot_results_added.rds")

# all_results[, data][[1]] %>%
#   ggplot() +  
#   geom_point(aes(y = yield, x = balance)) +
#   facet_grid(sat_cat ~ .)

### 2023-07-29 Writing/manuscript.Rmd
analysis updated

Two pathways exist by which aquifer depletion will impact farmers' ability to access and use groundwater as a buffer against drought and rainfall variability [@foster2015analysis]. First, depletion increases the energy requirements and costs of pumping water, potentially making it less economically viable for farmers to satisfy crop water needs [@mieno2017price; @bhattarai2021impact]. Second, depletion also reduces the pumping capacity of wells used to extract groundwater, which may create a physical constraint to the volume of irrigation that can be applied to a field [@konikow2005groundwater; @foster2014modeling; @hrozencik2017heterogeneous]. The combined effect of these two factors could force farmers to reduce irrigation depths, resulting in greater risk of drought-related crop stress. Alternatively, or in addition, a farmer may be forced to limit irrigated area to ensure crop water demands can be met, leading to reductions in total crop production [@foster2014modeling; @rad2020effects]. Understanding how these two responses vary as a function of aquifer conditions is crucial to guide groundwater management planning decisions, in particular given projected increases in frequency and intensity of droughts under climate change [@ukkola2020robust; @chiang2021evidence; @cook2020twenty]

@@

Figure \@ref(fig:ir-share) shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean. Figure \@ref(fig:ir-share) demonstrates clearly that aquifer thickness has large and significant impacts on the share of irrigated production, in particular for corn. As aquifer thickness declines, the share of irrigated acres for corn decreases from around 0.8 (80\%) for an average county-level aquifer thickness of 150m to 0.5 (50\%) for an average county-level aquifer thickness of 12m. Reductions in the share of production area under irrigation is also evident for soybean, although the effect is less pronounced than for corn potentially due to soybean's lower water requirements and susceptibility to drought than corn [@zipper2016drought; @ruess2022irrigation].

@@

## Aquifer depletion increases overall drought-related production losses by up to $25\%$

@@

res_at_950 <-
  all_results[, .(crop, avg_yield_pred_data)] %>%
  unnest(cols = c(avg_yield_pred_data)) %>%
  data.table() %>%
  .[(abs(balance - 950)) < 1, by = crop]
```
The overall effects of aquifer depletion on drought-related production losses is a function of both per-area irrigated yield and irrigated area changes highlighted in Sections \@ref(impact-intensive) and \@ref(impact-share). Combining both responses, Figure \@ref(fig:tot-impact) shows the relationship between water deficit and average per-area production (weighted-average of irrigated and rainfed yields, where weights are production area shares) for corn and soybean at thee values of aquifer thickness: 12.2, 59.1, 106.1, and 153.0  meters^[Confidence intervals were not presented to avoid over-crowding the figures. Please see Appendix \@ref(reg-conf) for estimated average yield response curves with 95 percent confidence interval.]. While you can see similar trends as presented in Figure \@ref(fig:yield-response), the difference in production loss caused by high water deficits among the aquifer thickness levels are much more pronounced mainly dues to lower irrigated area for lower aquifer thickness level (rainfed production has a larger share).

Figure \@ref(fig:dif-tot-impact) shows the difference in average per-area production of the 59.1m, 106.1m, and 153.0m aquifer thickness levels relative to 12.2m with the 95 percent confidence interval for corn and soybean. For corn, aquifer thickness does not have a significant impact on productivity when precipitation deficits are small (less than around 200mm and 300mm for 59.1m and 106.1m aquifer thickness, respectively). However, beyond this threshold, higher levels of water deficit result in noticeable and statistically significant reductions to average corn production. This reflects the combination of lower irrigated yields under drought conditions, and, more importantly, reduced overall productivity due to smaller share of land under irrigation in areas and years where aquifer thickness is lower. Similar effects are also observed for soybean, although reductions in average production only occur at a higher deficit threshold (i.e. greater than around 800mm) and are smaller in overall magnitude. For example, for a seasonal water deficit of 950mm (average water deficit in 2012, the most extreme drought in our historical record), average corn and soybean productivity are about 25\% and 15\% lower, respectively, when differentiating between areas with the highest (3rd quantile) and lowest (1st quantile) aquifer thickness. Further, notice that the decline in average per-area production at a seasonal water deficit of 950mm is not linear. Going from 153.0m aquifer thickness to 106.1m (decline of 46.9m) would result in `r round(res_at_950[sat == 502 & crop == "corn", avg_yield] - res_at_950[sat == 348 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to aquifer thickness when the aquifer thickness level is high.
Going from 106.1m aquifer thickness to 59.1m (decline of 47m) would result in `r round(res_at_950[sat == 348 & crop == "corn", avg_yield] - res_at_950[sat == 194 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn. However, going from 59.1m aquifer thickness to 12.2m (decline of 46.9m) would result in `r round(res_at_950[sat == 194 & crop == "corn", avg_yield] - res_at_950[sat == 40 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn. That is, the same degree of decline is much more damaging when the decline is from a low aquifer thickness level than a higher aquifer thickness level. This is primarily due to the non-linear impact of aquifer thickness on the share of irrigated corn acres as shown in Figure \@ref(fig:ir-share). When aquifer thickness level is high enough (e.g., 100m), a small decline does not reduce the share of irrigated acres. However, when aquifer thickness level is low (e.g., 40m), a small decline does reduce the share of irrigated acres.

@@

Several opportunities exist for future research to build upon the methods and analyses presented here. First, due to the limited resolution of our observational data, we were only able to estimate the impacts of different aquifer thickness categories on drought risk. Availability of sub-county crop yield data, for example from satellite remote sensing and/or in-situ monitoring [@edreira2020combining; @deines2021million], would allow models of the continuous effect of aquifer thickness on drought risk to be developed. This would enable more precise estimates to be made of critical tipping points in the resilience of irrigated agriculture to drought, which are needed to effectively design and target groundwater conservation policies.

Second, our analysis considered the effects of aquifer thickness on farmers decisions to switch from irrigated to rainfed production, but did not assess other potential responses to aquifer depletion such as changing crop types. Switching to more drought resistant crops or varieties may enable farmers to maintain irrigated production at lower levels of aquifer thickness, and has been shown to be an important response to increasing agricultural water scarcity and abstraction policies in other studies [@bhattarai2021impact; @deines2019quantifying; @manning2017producer]. Future research should therefore explore potential strategies that farmers could adopt to lessen the combined impacts of drought and aquifer depletion, alongside efforts to curb unsustainable groundwater extraction and climate change.

@@

To estimate the additional impact of aquifer thickness on the share of total acres that are irrigated in a county, a fractional logit model under the generalized additive model framework is used in which the impact of aquifer thickness, average water deficit, and their interactions are estimated without needing to assume any functional forms. This allows for flexible representation of the non-linear relationships between drought risks, aquifer conditions, and agricultural land use, including potential thresholds in farmers irrigated area choices under water supply constraints that have been noted in prior work (REF). The model is given in Equation \@ref(eq:eqn2) below:

>>>>

Two pathways exist by which aquifer depletion will impact farmers' ability to access and use groundwater as a buffer against drought and rainfall variability [@foster2015analysis]. First, depletion increases the energy requirements and costs of pumping water, potentially making it less economically viable for farmers to satisfy crop water needs [@mieno2017price; @bhattarai2021impact]. Second, depletion also reduces the aquifer thickness of an aquifer, lowering the aquifer's transmissivity (i.e., the potential rate of water supply to wells). This latter effect reduces the rate at which water can be pumped and applied to a field as irrigation [@konikow2005groundwater; @foster2014modeling; @hrozencik2017heterogeneous]. If the reduction in well yields is substantial - for example as would occur when a large proportion of aquifer thickness is depleted [@hecox2002calculation; @korus2020depletion] - then a farmer may be unable to extract sufficient water to meet crop water demands, in particular during periods of limited rainfall and in the peak of the growing season when crops are most sensitive to water deficits \citep{foster2015well,rouhi2020downside}. The combined effect of these two factors could force farmers to reduce irrigation depths, resulting in greater risk of drought-related crop stress. Alternatively, or in addition, a farmer may be forced to limit irrigated area to ensure crop water demands can be met, leading to reductions in total crop production [@foster2014modeling; @rad2020effects]. Understanding how these two responses vary as a function of aquifer conditions is crucial to guide groundwater management planning decisions, in particular given projected increases in frequency and intensity of droughts under climate change [@ukkola2020robust; @chiang2021evidence; @cook2020twenty]

@@

corn_share <- all_results[crop == "corn", share_pred_data][[1]]

share_high <- 
  corn_share[which.min(abs(sat - 150)), ir_share_hat] %>%
  round(digits = 2)

share_low <- 
  corn_share[which.min(abs(sat - 10)), ir_share_hat] %>%
  round(digits = 2)
```

Figure \@ref(fig:ir-share) shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in drought risk exposure across the aquifer. Figure \@ref(fig:ir-share) demonstrates clearly that aquifer thickness has large and significant impacts on the share of irrigated production, in particular for corn. As aquifer thickness declines, the share of irrigated acres for corn decreases from around `r share_high` (`r share_high * 100`\%) for an average county-level aquifer thickness of 150m to `r share_low` (`r share_low * 100`\%) for an average county-level aquifer thickness of 12m. Reductions in the share of production area under irrigation is also evident for soybean, although the effect is less pronounced than for corn potentially due to soybean's lower water requirements and susceptibility to drought than corn [@zipper2016drought; @ruess2022irrigation].

@@

## Aquifer depletion increases overall drought-related production losses by up to $25\%$j

@@

res_at_950 <-
  all_results[, .(crop, avg_yield_pred_data)] %>%
  unnest(cols = c(avg_yield_pred_data)) %>%
  data.table() %>%
  .[crop == "corn", ] %>%
  .[(abs(balance - 950)) < 5, ]
```

The overall effects of aquifer depletion on drought-related production losses is a function of both per-area irrigated yield and irrigated area changes highlighted in Sections \@ref(impact-intensive) and \@ref(impact-share). Combining both responses, Figure \@ref(fig:tot-impact) shows the relationship between water deficit and average per-area production (weighted-average of irrigated and rainfed yields, where weights are production area shares) for corn and soybean at thee values of aquifer thickness: 10, 50, 90, and 130 meters^[Confidence intervals were not presented to avoid over-crowding the figures. Please see Appendix \@ref(reg-conf) for estimated average yield response curves with 95 percent confidence interval.]. While you can see similar trends as presented in Figure \@ref(fig:yield-response), the difference in production loss caused by high water deficits among the aquifer thickness levels are much more pronounced mainly dues to lower irrigated area for lower aquifer thickness level (rainfed production has a larger share).

Figure \@ref(fig:dif-tot-impact) shows the difference in average per-area production of the 50m, 90m, and 130m aquifer thickness levels relative to 10m with the 95 percent confidence interval for corn and soybean. For corn, aquifer thickness does not have a significant impact on productivity when precipitation deficits are small (less than around 200mm and 300mm for 50m and 90m aquifer thickness, respectively). However, beyond this threshold, higher levels of water deficit result in noticeable and statistically significant reductions to average corn production. This reflects the combination of lower irrigated yields under drought conditions, and, more importantly, reduced overall productivity due to smaller share of land under irrigation in areas and years where aquifer thickness is lower. Similar effects are also observed for soybean, although reductions in average production only occur at a higher deficit threshold (i.e. greater than around 800mm) and are smaller in overall magnitude. For example, for a seasonal water deficit of 950mm (average water deficit in 2012, the most extreme drought in our historical record), average corn and soybean productivity are about 25\% and 15\% lower, respectively, when differentiating between areas with the highest (3rd quantile) and lowest (1st quantile) aquifer thickness.

Further, notice that the impact of aquifer thickness on the decline in average per-area production is highly non-linear at high seasonal water deficits. Going from 130m aquifer thickness to 90m (decline of 40m) would result in merely `r round(res_at_950[sat == 130 & crop == "corn", avg_yield] - res_at_950[sat == 90 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to aquifer thickness when the aquifer thickness level is high.
Going from 90m aquifer thickness to 50m (decline of 40m) would result in `r round(res_at_950[sat == 90 & crop == "corn", avg_yield] - res_at_950[sat == 50 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn. However, going from 50m aquifer thickness to 10m (decline of 40m) would result in `r round(res_at_950[sat == 50 & crop == "corn", avg_yield] - res_at_950[sat == 10 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn. That is, the same degree of decline is much more damaging when the decline is from a low aquifer thickness level than a higher aquifer thickness level. This is primarily due to the non-linear impact of aquifer thickness on the share of irrigated corn acres as shown in Figure \@ref(fig:ir-share). When aquifer thickness level is high enough (e.g., 100m), a small decline does not reduce the share of irrigated acres. However, when aquifer thickness level is low (e.g., 40m), a small decline does reduce the share of irrigated acres.

@@

Several opportunities exist for future research to build upon the methods and analyses presented here. First, due to the limited resolution of our observational data, we were only able to estimate the impacts of different aquifer thickness categories on drought risk. Availability of sub-county crop yield data, for example from satellite remote sensing and/or in-situ monitoring [@edreira2020combining; @deines2021million], would allow models of the continuous effect of aquifer thickness on drought risk to be developed. Similarly, unobserved variations in aquifer properties will also introduce noise into the relationships between aquifer thickness, well yields, and drought risk exposure. For example, a thinner aquifer interval composed of coarse sands and gravels may be able to supply comparable or higher well yields than a thicker aquifer dominated by fine sands and silt deposits, and hence offer greater resilience to drought despite similar or larger aquifer thickness [@butler2013interpretation; @korus2020depletion]. While we identify statistically robust impacts of aquifer thickness on drought risk despite these uncertainties, improved data on spatial variations in aquifer properties and greater monitoring of real-world changes in well capacities alongside traditional water level measurements would help to refine understanding of patterns of drought vulnerability across the HPA. This would enable more precise estimates to be made of critical tipping points in the resilience of irrigated agriculture to drought, which are needed to effectively design and target groundwater conservation policies.

Second, our analysis considered the effects of aquifer thickness on farmers decisions to switch from irrigated to rainfed production, but did not assess other potential responses to aquifer depletion such as changing crop types. Switching to more drought resistant crops or varieties may enable farmers to maintain irrigated production at lower levels of aquifer thickness, and has been shown to be an important response to increasing agricultural water scarcity and abstraction policies in other studies [@bhattarai2021impact; @deines2019quantifying; @manning2017producer]. Improvements to irrigation scheduling and water management practices, such as have been implemented as part of the SD6-LEMA (Local Enhanced Management Area) in Kansas [@deines2019quantifying; @glose2022quantifying] and proposed as part of California's SGMA (Sustainable Groundwater Management Act) [@berbel2019droughts; @lubell2020sustainable], could also play an important role in improving the efficiency of irrigation and, hence, the ability to more effectively meet crop water demands with lower well capacities. These kinds of interventions would be likely to reduce - but not eliminate - the increases in drought risk caused by declining aquifer thickness, and could bring additional benefits of reducing over-use of water in wetter years through greater awareness of efficiency saving potentials [@foster2019assessing] and the importance of conservation [@marston2022] amongst growers. Future research should therefore explore potential strategies that farmers could adopt to lessen the combined impacts of drought and aquifer depletion, alongside efforts to curb unsustainable groundwater extraction and climate change.

@@

To estimate the additional impact of aquifer thickness on the share of total acres that are irrigated in a county, a fractional logit model under the generalized additive model framework is used in which the impact of aquifer thickness, average water deficit, and their interactions are estimated without needing to assume any functional forms. This allows for flexible representation of the non-linear relationships between drought risks, aquifer conditions, and agricultural land use, including potential thresholds in farmers irrigated area choices under water supply constraints that have been noted in prior work [@foster2015well; @foster2015analysis; @hrozencik2017heterogeneous]. The model is given in Equation \@ref(eq:eqn2) below:
